---
title: "NYC Shooting Data Analysis DTSA_5301"
author: "R. Dadmun"
date: "2024-01-22"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This is a breakdown of every shooting incident that occurred in NYC from 2006-2023. This data is manually extracted every quarter and reviewed by the Office of Management Analysis and Planning before being posted on the NYPD website. Each record represents a shooting incident in NYC and includes information about the event, the location and time of occurrence. Please refer to [NYPD Shooting Incident Data (Historic) - CKAN] https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic for more information about this dataset.

## Purpose

The purpose of this document is to show that I can create an Rmarkdown document to highlight my data analysis process, and provide a guide to reproduce the analysis I performed for DTSA 5301 - Data Science as a Field at the University of Colorado Boulder. This analysis will investigate which age groups, genders and race demographics are most likely to be the victim of a shooting incident in NYC, and which boroughs are most likely to have a shooting incident occur. Finally, we will develop a basic regression model to see if any statistic about the victim can be used to determine if the shooting will be fatal or not.  

### Import Libraries

The following code will install the tidyverse, lubridate, and the ggthemes library to brighten up our visualizations, and the second line will include it as a library within the Rmarkdown file:

```{r libraries, message=FALSE, warning=FALSE}
#install.packages("tidyverse")
#install.packages("lubridate")
#install.packages("ggthemes")
library(tidyverse)
library(lubridate)
library(ggthemes)
```

Next, we would like to ensure we have a connection point to our dataset. This dataset was pulled from <https://catalog.data.gov> as a CSV. We utilize the "read.csv" command to ensure our dataset is imported into R.

```{r import_NYC_Shooting_Data}
NYC_Shootings <- read.csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD", header = TRUE)
head(NYC_Shootings)
```

## Cleaning the data

After the data has been loaded in, we want to take the time to clean the data. We have removed 9 fields from the data in order to slim the dataset down. 

Removed Fields:
 **PRECINCT**,
 **JURISDICTION_CODE**,
 **LOCATION_DESC**,
 **X_COORD_CD**, 
 **Y_COORD_CD**,
 **Lat**,
 **Long**,
and **Lon_Lat**. 

```{r Delete Unused Fields}
NYC_Shootings_2 <- NYC_Shootings %>%
  select(INCIDENT_KEY,
  OCCUR_DATE,
  OCCUR_TIME,
  BORO,
  STATISTICAL_MURDER_FLAG,
  PERP_AGE_GROUP,
  PERP_SEX,
  PERP_RACE,
  VIC_AGE_GROUP,
  VIC_SEX,
  VIC_RACE)

summary(NYC_Shootings_2)
```
### Observations
Key observations on data type conversion are:

* **INCIDENT_KEY** should be treated as a string.
* **BORO** should be treated as a factor.
* **PERP_AGE_GROUP** should be treated as a factor.
* **PERP_SEX** should be treated as a factor.
* **PERP_RACE** should be treated as a factor.
* **VIC_AGE_GROUP** should be treated as a factor.
* **VIC_SEX** should be treated as a factor.
* **VIC_RACE** should be treated as a factor.

We also ensure our NA factors are removed from the dataset, and every remaining Unknown variable matches the same format. There is a single value in the **VIC_AGE_GROUP** field marked as 1022, which we will move and merge with Unknown.

```{r Data Tidy and Transformations}
# Remove NA Values in data
NYC_Shootings_2 <- NYC_Shootings_2 %>%
  replace_na(list(PERP_AGE_GROUP = "Unknown", PERP_SEX = "Unknown", PERP_RACE = "Unknown"))

#Change the coding and element types per field
NYC_Shootings_2$VIC_SEX = recode(NYC_Shootings_2$VIC_SEX, U = "Unknown")
NYC_Shootings_2$VIC_RACE = recode(NYC_Shootings_2$VIC_RACE, UNKNOWN = "Unknown")
NYC_Shootings_2$VIC_AGE_GROUP = recode(NYC_Shootings_2$VIC_AGE_GROUP, "1022" = "Unknown")
NYC_Shootings_2$VIC_AGE_GROUP = recode(NYC_Shootings_2$VIC_AGE_GROUP, UNKNOWN = "Unknown")
NYC_Shootings_2$INCIDENT_KEY = as.character(NYC_Shootings_2$INCIDENT_KEY)
NYC_Shootings_2$BORO = as.factor(NYC_Shootings_2$BORO)
NYC_Shootings_2$PERP_AGE_GROUP = as.factor(NYC_Shootings_2$PERP_AGE_GROUP)
NYC_Shootings_2$PERP_SEX = as.factor(NYC_Shootings_2$PERP_SEX)
NYC_Shootings_2$PERP_RACE = as.factor(NYC_Shootings_2$PERP_RACE)
NYC_Shootings_2$VIC_AGE_GROUP = as.factor(NYC_Shootings_2$VIC_AGE_GROUP)
NYC_Shootings_2$VIC_SEX = as.factor(NYC_Shootings_2$VIC_SEX)
NYC_Shootings_2$VIC_RACE = as.factor(NYC_Shootings_2$VIC_RACE)
NYC_Shootings_2$STATISTICAL_MURDER_FLAG <- factor(NYC_Shootings_2$STATISTICAL_MURDER_FLAG)

# Return summary statistics
summary(NYC_Shootings_2)
```
## Checking for additional missing data:

We will locate where data is missing within the data using the colSums function. Next, we will check the total number of missing data elements (which are marked as NA), and the percent of the data which is missing:

```{r Missing Data Check}
colSums(is.na(NYC_Shootings_2))

```
As we can see, our data is clean with no NA entries.

## Research Questions:

1. Which part of New York has the most number of incidents? Of those incidents, how many are murder cases? 

2. Which groups of people were most likely to be victims of a shooting in NYC?

## Visualizing and Analyzing the Data

### 1. Which part of New York has the most number of incidents? Of those incidents, how many are murder cases? 

```{r Q1}
Boro_Graph <- ggplot(NYC_Shootings_2, aes(x = BORO)) +
  geom_bar(fill = "blue") +
  labs(title = "Shooting Incidents in NYC by Borough of Occurence",
       x = "Boroughs of New York City",
       y = "Count of Incidents") +
  theme_bw()

table(NYC_Shootings_2$BORO, NYC_Shootings_2$STATISTICAL_MURDER_FLAG)
table(NYC_Shootings_2$BORO)

Boro_Graph
```
Based on the graph, we can see that Brooklyn is 1st in the number of incidents, with 10,932 incidents, 2,122 of which were flagged as murders. The Bronx is significantly lower than this, with 7,935 incidents, of which 1,542 were flagged as murders. Queens then follows, with 4,094 incidents. Both Manhattan and Staten Island both have a significantly smaller number of incidents than Brooklyn, with Staten Island having both their number of incidents and the flags for murder numbering in the triple digits. 

### 2. Which groups of people were most likely to be victims of a fatal shooting in NYC?

#### Which age group is most likely to be the victim of a shooting incident in NYC?

First, we want to see how many of our shootings were fatal, which is summarized in the STATISTICAL_MURDER_FLAG field:

```{r}
Fatal_Tbl <- table(NYC_Shootings_2$STATISTICAL_MURDER_FLAG)
Fatal_Tbl
```

At the time of this analysis, we can see that there are 22,046 non-fatal shootings in the dataset, and 5,266 fatal shootings.

Next, we would like to check how our fatal shootings are distributed by victim age:

```{r Age_x_Fatal}
Fatal_Age_Tbl <- table(NYC_Shootings_2$STATISTICAL_MURDER_FLAG, NYC_Shootings_2$VIC_AGE_GROUP)
Fatal_Age_Tbl
```
From the table above, we can see that the vast majority of victims of shooting incidents were in the 18-24 and 25-44 age groups. 
We will utilize a bar chart to compare this distributions in a more visual way:

```{r Fatal_Age_Bar}
Age_Graph <- NYC_Shootings_2 %>%
  filter(VIC_AGE_GROUP != "Unknown") %>%
  ggplot(aes(x = VIC_AGE_GROUP, fill = (STATISTICAL_MURDER_FLAG))) +
  geom_bar(position = "dodge") +
  theme_bw() +
  labs(title = "NYC Shooting Incidents by Age Group",
      x = "Victim Age Group",
      y = "Total Shooting Incidents",
      fill = "Flagged as Murder")

Age_Graph
```
Based on the table above, I hypothesize that age group is correlated with the STATISTICAL_MURDER_FLAG field. 

#### Which gender is more likely to be a victim of a shooting incident in NYC? 

Next, we will run a quick analysis on which gender is more likely toi be the victim of a shooting incident in NYC. In our data, genbder is marked by the VIC_SEX field, and is given values of Male, Female or Unknown. We will make a table to list out the exact count of each, and then create a graph to better visualize this data. The graph will also show the value differences between fatal and non-fatal shooting incidents. 

```{r Gender_Tbl}
table(NYC_Shootings_2$VIC_SEX)
```

```{r Gender Graph}
Gender_Graph_Stack <- NYC_Shootings_2 %>%
  filter(VIC_SEX != "Unknown") %>%
  ggplot(aes(x = VIC_SEX, fill = (STATISTICAL_MURDER_FLAG))) +
  geom_bar() +
  theme_bw() +
  labs(title = "NYC Shooting Incidents by Sex",
      x = "Victim Sex",
      y = "Total Shooting Incidents",
      fill = "Flagged as Murder")

Gender_Graph_Stack
```
Based on the above graphs, we can see that Men are unilaterally more likely to be the victim of a shooting even in NYC than Women. 

#### Which demographic race is most likely to be the victim of a shooting event in NYC?

Finally, we would like to explore the rates of shooting incidents in NYC by victim race. 

``` {r}
table(NYC_Shootings_2$VIC_RACE)

Race_Graph <- NYC_Shootings_2 %>%
  filter(VIC_RACE != "Unknown") %>%
  ggplot(aes(y = VIC_RACE, fill = (STATISTICAL_MURDER_FLAG))) +
  geom_bar(position = "dodge") +
  theme_bw() +
  labs(title = "NYC Shooting Incidents by Race",
      x = "Victim Sex",
      y = "Total Shooting Incidents",
      fill = "Flagged as Murder")

Race_Graph
```
Based on the graphs above, we have a few major takeaways to flag for future analysis:  
  1. Brooklyn and the Bronx are by far the most likely boroughs in which a shooting incident may occur.   
  2. Men are far more likely to be the victim of a shooting incident than women.  
  3. Black individuals are at a much higher likelihood to be the victim of a shooting incident in NYC. Both White Hispanic and Black Hispanic individuals follow in second and third place as the most likely demographic to be the victim of a shooting event.

Summarized in a sentence: Black or Hispanic males in Brooklyn or the Bronx have an elevated chance to be the victim of a shooting event as opposed to other population demographics or locations in NYC

# Multivariable Logistic Regression Model
Logistic regression models work well with a plethora of categorical variables, of which this data set has many. In our model, we will be utilizing the victim's age, gender and race. 

The objective of the model is to determine if any of the aforementioned variables can be used to predict if a shooting will be fatal or not.

**Independent Variable** STATISTICAL_MURDER_FLAG
  This variable indicates whether a shooting was fatal or not. True (1) signals that the shooting was fatal, while False (0) indicates that the shooting was not fatal. 
**Dependent Variable** VIC_AGE_GROUP, VIC_RACE, VIC_SEX

```{r Regression Model}
Reg_Model <- glm(STATISTICAL_MURDER_FLAG ~ VIC_AGE_GROUP + VIC_SEX + VIC_RACE, data = NYC_Shootings_2, family = 'binomial')

summary(Reg_Model)

```

**Observations:**
The **victim's age group** seems to be the most significant variable for determining how likely a victim is to survive a shooting incident in NYC. More specifically, a victim is **most likely to survive the incident** if they are in the **< 18 or 18-24 age groups**. As the age increases, the liklihood of survival diminished. With a coefficient greater than 1 for the **65+ age group**, it appears that most shooting events in this age group are fatal.

Both **Victim Age and Victim Race** appear to be uncorrelated in the determination of whether a shooting event in NYC is fatal to the victim. No singular race demographic had a significant correlation with the determination as to if shooting incidents in NYC ended in murder; nor did the sex of the victim have aa significant correlation value to whether or not a murder occured.

## Identifing Bias
The topics of gun violence and gender are both sources of bias for me. I assumed initially that women would more likely be the incident of crime, and tried to ensure I approached this topic as neutrally as possible to ensure I was ready for the takeaways the data showed me. Additionally, my political stances on gun ownership and gun violence would have lead me to believe most gun crimes end in death. Overall, I found it relitively easy to remain objective in my analysis as there was little context or discussion surrounding this dataset prior to exploration. 


##Resources
- https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic  
- https://data.cityofnewyork.us/Public-Safety/NYPD-Shooting-Incident-Data-Historic-/833y-fsy8/about_data  
- https://www.vitalcitynyc.org/vital_signs/gun-violence-in-new-york-city-the-data  
- https://uc-r.github.io/missing_values   
- https://www.codecademy.com/learn/learn-linear-regression-in-r/modules/linear-regression-in-r/cheatsheet   
- https://r4ds.had.co.nz/index.html  
